Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-11-23 FreeLogic/2018-JAR/article.tex
update

\author{Christoph Benzm\"uller  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   Freie Universit\"at Berlin, Berlin, Germany \& University of Luxembourg, Luxembourg\\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

\begin{defn}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subset D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. $\True$ or $\False$ denote truth and
  falsehood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{defn}

@@

The sources of our experiments as conducted here are available at
\url{http://christoph-benzmueller.de/papers/2018-JAR-sources.zip}.
These sources contain an embedding of full free logic in Isabelle/HOL,
that is, with $\star$ and definite description
(cf.~Fig.\ref{pic1}). In \cite{ArXiv}, $\star$ and definite
description were still avoided; they are in fact not really 
relevant for the conducted experiments.
% Note that
% in these sources we added neither the definite description operator
% nor the designated undefined object $\hol{\star}$, since both were not
% required in this particular application. A related Isabelle/HOL
% encoding of the above theory with description has been presented and
% discussed in \cite{ICMS}.

@@

\subsection{Consistency}
The model finder Nitpick confirms consistency for all of the axioms
sets from Table \ref{axioms-sets-1}. For example, when asked to consider at least one defined and one undefined object, then
Nitpick generates for all cases the following model (called $M_1$ in the remainder):
$D=\{i_i,i_2\}$ and $E=\{i_1\}$;  $i_1\comp i_1$ is $i_1$,  and $i_2$
in all other cases; $cod$ and $dom$ are identity on $D$. Without constraining the request, Nitpick
generates an even simpler model (called $M_0$ in the remainder):
$D=\{i_i\}$ and $E=\emptyset$; $i_1\comp i_1$ is $i_1$; $cod$ and
$dom$ are identity on $D$. It is trivial to check that these models
indeed confirm the consistency of all axioms sets from Table \ref{axioms-sets-1}.

@@

  In our experiments we have also made use of the Isabelle/HOL's
  \textsc{smt} method, which ``translates the conjecture and any
  user-supplied facts to the SMT solvers’ many-sorted first-order
  logic, invokes a solver, and (depending on the solver) either trusts
  the result or attempts to reconstruct the proof in Isabelle.''
  (\cite[p.~5]{Sledgehammer}).\footnote{Technical remark: We have
    selected CVC4 in our experiments as the default SMT solver, since
    we did run into errors when working with Z3. These errors can
    easily be reconstructed in the provided source files when
    switching back to Z3 as default.}  For quite some time the use of the \textsc{smt}
  method has been controversially discussed in the Isabelle/HOL
  community. However, since about a year, the \textsc{smt} method is
  now even eligible in submissions to the Archive of Formal
  Proofs. For the work presented here this community internal discussion is 
  of little relevance, since we could replace all uses of the \textsc{smt} method
  by alternative proofs (which we avoid in order to keep
  our formalisation concise).

@@

 \section{Assessment of the Axiom System by Freyd and
   Scedrov} \label{sec-freyd-scedrov} In this section we study the
 axioms set of Freyd and Scedrov from their textbook ``Categories,
 Allegories'' \cite{FreydScedrov90}.  In \S\ref{subsec-freyd-scedrov-1} we show, that their axioms set,
 replicated in Table \ref{axioms-sets-2} as Axioms Set FS-I, becomes
 inconsistent in our free logic setting if we assume non-existing
 objects in $D$, respectively, if we assume that the operations
 are non-total.

@@

 \subsection{Constricted Inconsistency in Free Logic
   Setting} \label{subsec-freyd-scedrov-1} A main difference in the
 system by Freyd and Scedrov to our Axioms Set V from Table
 \ref{axioms-sets-1} concerns axiom $S3$, respectively $A1$. Namely,
 instead of the non-reflexive existing identity $\simeq$, they use
 Kleene equality $\cong$, cf. definition 1.11 on page 3 of their textbook
 \cite{FreydScedrov90}.\footnote{Def. 1.11 in Freyd Scedrov: ``The
   ordinary equality sign $=$ [i.e., our $\cong$] will be used in the
   symmetric sense, to wit: if either side is defined then so is the
   other and they are equal. \ldots''} The difference seems minor, but
 in our free logic setting it has the effect to cause the mentioned
 constricted inconsistency issue.\footnote{This could perhaps be an
   oversight, or it could indicate that Freyd and Scedrov actually
   mean the Axioms Set discussed in \S\ref{subsec-freyd-scedrov-2}
   below.%  (where the variables in the axioms range over defined objects
   % only). However, in this reading of the axioms of Freyd and Scedrov
   % we had to (re-)introduce explicit strictness conditions to ensure
   % equivalence to the Axioms Set V by Scott.
 }

  The (constricted) inconsistency of Axioms Set FS-I, respectively
  Axioms Set FS-II, from Table \ref{axioms-sets-2} has been 
  detected first by the model finder Nitpick. When we asked Nitpick to
  generate a model with at least one non-existing object, it claimed
  that there is no such model. However, a model can still be
  constructed if we do not make any assumptions about non-existing
  objects.\footnote{For this we have to inactivate the axiom
    that postulates that $\star$ is an undefined/non-existing object.} In fact, the model presented by Nitpick for this case
  consists of a single, existing morphism.

@@

For Axioms Set FS-III the consistency checks with Nitpick succeed,
even if we assume undefined objects. However, this axioms set is
obviously weaker than Axioms Set V from Table
\ref{axioms-sets-1}. In fact, as has been shown by Nitpick, none of
the axioms of this set are implied.  The situation changes when we
explicitly postulate strictness of $dom$, $cod$ and $\comp$. Doing
so we obtain Axioms Set FS-IV from Table \ref{axioms-sets-4}, which,
as Nitpick confirms, is consistent even if we assume undefined
objects. And the automated theorem provers via Sledgehammer confirm
that Axioms Set FS-IV is equivalent to Axioms Set V, as
intended. Unfortunately, however, respective strictness conditions are
not mentioned in the textbook by Freyd and Scedrov.

@@

\begin{acknowledgements}
This work has been supported by the DFG Heisenberg fellowship (BE
2501/9-2, \emph{Towards Computational Metaphysics}) of the first
author.

We thank G{\"u}nter Rote, Lutz Schr\"oder and and Emil Weydert for their
comments to \cite{ArXiv}, which together with \cite{ICMS} forms the
basis for this article.

\end{acknowledgements}

>>>>

\author{Christoph Benzm\"uller\thanks{Benzmüller received funding from the German National Research
  Foundation DFG under Heisenberg grant \emph{Towards Computational
    Metaphysics} (BE 2501/9-2) and from VolkswagenStiftung under grant
  Consistent Rational Argumentation in Politics (CRAP).}  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   Freie Universit\"at Berlin, Berlin, Germany \& University of Luxembourg, Luxembourg\\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

\begin{defn}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subset D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \dotsb \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \dotsb \times D$. $\True$ or $\False$ denote truth and
  falsehood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{defn}

@@

The sources of our experiments as conducted here are available at
\url{http://christoph-benzmueller.de/papers/2018-JAR-sources.zip}.
These sources contain an embedding of full free logic in Isabelle/HOL,
that is, with $\star$ and definite description
(cf.~Fig.~\ref{pic1}). In \cite{ArXiv}, $\star$ and definite
description were still avoided; they are in fact not really 
relevant for the conducted experiments.
% Note that
% in these sources we added neither the definite description operator
% nor the designated undefined object $\hol{\star}$, since both were not
% required in this particular application. A related Isabelle/HOL
% encoding of the above theory with description has been presented and
% discussed in \cite{ICMS}.

@@

\subsection{Consistency}
The model finder Nitpick confirms consistency for all of the axioms
sets from Table~\ref{axioms-sets-1}. For example, when asked to consider at least one defined and one undefined object, then
Nitpick generates for all cases the following model (called $M_1$ in the remainder):
$D=\{i_i,i_2\}$ and $E=\{i_1\}$;  $i_1\comp i_1$ is $i_1$,  and $i_2$
in all other cases; $cod$ and $dom$ are identity on $D$. Without constraining the request, Nitpick
generates an even simpler model (called $M_0$ in the remainder):
$D=\{i_i\}$ and $E=\emptyset$; $i_1\comp i_1$ is $i_1$; $cod$ and
$dom$ are identity on $D$. It is trivial to check that these models
indeed confirm the consistency of all axioms sets from Table~\ref{axioms-sets-1}.

@@

  In our experiments we have also made use of the Isabelle/HOL's
  \textsc{smt} method, which ``translates the conjecture and any
  user-supplied facts to the SMT solvers’ many-sorted first-order
  logic, invokes a solver, and (depending on the solver) either trusts
  the result or attempts to reconstruct the proof in Isabelle.''
  \cite[p.~5]{Sledgehammer}.\footnote{Technical remark: We have
    selected CVC4 in our experiments as the default SMT solver, since
    we did run into errors when working with Z3. These errors can
    easily be reconstructed in the provided source files when
    switching back to Z3 as default.}  For quite some time the use of
  the \textsc{smt} method has been controversially discussed in the
  Isabelle/HOL community, and there is in fact a significant difference
  between using the \textsc{smt} method in combination with Z3 or with
  CVC4, as we prefer. When setting the solver to CVC4, the contributed
  proofs are accepted and being trusted without replaying them in the
  Isabelle kernel. Proofs contributed by Z3, in contrast, are never
  trusted and always replayed in Isabelle's kernel. For the work
  presented here this community internal discussion is of minor
  relevance, so that we decided to continue working with CVC4 in order
  to keep our formalisation concise and also because CVC4
  performed surprisingly well in our experiments.\footnote{An expert
    reviewer of this article, to whom we are very grateful, provided
    alternative proofs which can be
    fully replayed in the kernel of Isabelle.}

@@

 \section{Assessment of the Axiom System by Freyd and
   Scedrov} \label{sec-freyd-scedrov} In this section we study the
 axioms set of Freyd and Scedrov from their textbook ``Categories,
 Allegories'' \cite{FreydScedrov90}.  In \S\ref{subsec-freyd-scedrov-1} we show that their axioms set,
 replicated in Table~\ref{axioms-sets-2} as Axioms Set FS-I, becomes
 inconsistent in our free logic setting if we assume non-existing
 objects in $D$, respectively, if we assume that the operations
 are non-total.

@@

 \subsection{Constricted Inconsistency in Free Logic
   Setting} \label{subsec-freyd-scedrov-1} A main difference in the
 system by Freyd and Scedrov to our Axioms Set V from Table~\ref{axioms-sets-1} concerns axiom $S3$, respectively $A1$. Namely,
 instead of the non-reflexive existing identity $\simeq$, they use
 Kleene equality $\cong$, cf. definition 1.11 on page 3 of their textbook
 \cite{FreydScedrov90}.\footnote{Def. 1.11 in Freyd Scedrov: ``The
   ordinary equality sign $=$ [i.e., our $\cong$] will be used in the
   symmetric sense, to wit: if either side is defined then so is the
   other and they are equal. \ldots''} The difference seems minor, but
 in our free logic setting it has the effect to cause the mentioned
 constricted inconsistency issue.\footnote{This could perhaps be an
   oversight, or it could indicate that Freyd and Scedrov actually
   mean the Axioms Set discussed in \S\ref{subsec-freyd-scedrov-2}
   below.%  (where the variables in the axioms range over defined objects
   % only). However, in this reading of the axioms of Freyd and Scedrov
   % we had to (re-)introduce explicit strictness conditions to ensure
   % equivalence to the Axioms Set V by Scott.
 }

  The (constricted) inconsistency of Axioms Set FS-I, respectively
  Axioms Set FS-II, from Table~\ref{axioms-sets-2} has been 
  detected first by the model finder Nitpick. When we asked Nitpick to
  generate a model with at least one non-existing object, it claimed
  that there is no such model. However, a model can still be
  constructed if we do not make any assumptions about non-existing
  objects.\footnote{For this we have to inactivate the axiom
    that postulates that $\star$ is an undefined/non-existing object.} In fact, the model presented by Nitpick for this case
  consists of a single, existing morphism.

@@

For Axioms Set FS-III the consistency checks with Nitpick succeed,
even if we assume undefined objects. However, this axioms set is
obviously weaker than Axioms Set V from Table~\ref{axioms-sets-1}. In fact, as has been shown by Nitpick, none of
the axioms of this set are implied.  The situation changes when we
explicitly postulate strictness of $dom$, $cod$ and $\comp$. Doing
so we obtain Axioms Set FS-IV from Table~\ref{axioms-sets-4}, which,
as Nitpick confirms, is consistent even if we assume undefined
objects. And the automated theorem provers via Sledgehammer confirm
that Axioms Set FS-IV is equivalent to Axioms Set V, as
intended. Unfortunately, however, respective strictness conditions are
not mentioned in the textbook by Freyd and Scedrov.

@@

\begin{acknowledgements}
 We thank G{\"u}nter Rote, Lutz Schr\"oder and and Emil Weydert for their
comments to \cite{ArXiv}, which together with \cite{ICMS} forms the
basis for this article.

We also want to express our gratitude to the reviewers of this
article.  Their fruitful feedback definitely helped to improve the
final version.
\end{acknowledgements}
